---
title: " Climate-Livestock Dynamics in Somalia"
subtitle: "Yale University | Environmental Data Science Capstone"
author: "Mohamed Said"
date: "2026-02-01"
output:
  html_document:
    theme: lumen            # Professional clean theme
    highlight: tango        # Nice code syntax highlighting
    toc: true               # Table of Contents
    toc_float: true         # TOC stays on the side as you scroll
    code_folding: hide      # Hides code by default, adds "Show/Hide" buttons
    number_sections: true   # Auto-numbers your headers
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
options(repos = c(CRAN = "https://cloud.r-project.org"))

```

# Abstract

Pastoral livestock systems in Somalia are highly dependent on rainfall because precipitation controls pasture availability, water resources, and rangeland productivity. However, rainfall in Somalia is characterized by strong interannual variability and frequent droughts, raising concerns about how climate fluctuations may influence livestock population dynamics. This study investigates the relationship between long-term rainfall variability and livestock stocks in Somalia from 1981 to 2024.

The analysis integrates satellite-derived rainfall data from the Climate Hazards Group InfraRed Precipitation with Station Data (CHIRPS v3) with national livestock stock statistics from the FAOSTAT database. Rainfall rasters were processed using spatial methods to generate national and regional rainfall summaries, rainfall anomalies, and spatial variability measures. Exploratory data analysis was used to examine rainfall distributions, drought years, and spatial rainfall patterns across Somalia’s administrative regions. Livestock population trends for camels, cattle, goats, and sheep were evaluated using time-series indicators including year-over-year growth rates and indexed stock levels. To evaluate potential climate–livestock linkages, a generalized additive model (GAM) was estimated using national rainfall anomalies and lagged rainfall anomalies (0–2 years) as predictors of log-transformed livestock stocks.

Results show that Somalia’s rainfall regime is dominated by strong year-to-year variability rather than a significant long-term trend. Livestock trends differ by species, with camels showing a significant long-term increase, goats showing a substantial decline, cattle remaining relatively stable, and sheep displaying weaker downward tendencies. The GAM analysis indicates that species-specific long-term time dynamics explain most variation in livestock stocks, while national rainfall anomalies and their lagged effects provide limited additional explanatory power.

These findings suggest that livestock dynamics in Somalia are shaped by multiple interacting drivers beyond national rainfall variability alone, including socio-economic conditions, conflict, market forces, and pastoral mobility constraints. Improving climate resilience in Somalia’s pastoral systems therefore requires integrated monitoring of climate, livestock, and socio-economic factors.

# Introduction

Pastoral livestock production is central to Somalia’s economy and rural livelihoods.Livestock contribute significantly to household income, food security, and national exports.

However, pastoral systems depend heavily on rainfall because precipitation determines 
rangeland productivity, pasture availability, and water access for livestock.

Somalia experiences highly variable rainfall with frequent droughts and occasional  extreme wet years. These climate fluctuations may affect livestock survival, herd growth, and long-term livestock population dynamics. Understanding how rainfall variability relates to livestock stocks is therefore important for climate adaptation planning and sustainable pastoral management.

*Research question How do long-term rainfall trends influence density changes of goats, sheep, and camels across Somalia's?.*


```{r,  eval=FALSE, include=FALSE}
## Install packages
install.packages("terra")          # raster data for rainfal
install.packages("sf")             # vector data for shapefiles
install.packages("dplyr")          # data cleaning and transformation
install.packages("tidyr")          # data reshaping and structuring
install.packages("exactextractr")  # zonal statistics by livelihood zones
install.packages("units")          # unit handling
install.packages("readr")          # reading and writing tables
install.packages("ggplot2")        # for nice graphs and visualizations
install.packages("viridis")       # smooth gradient collr for maps
install.packages("tidyterra")     # ggplot2 geoms for terra objects
install.packages(c("Kendall","trend")) # trend and Mann-Kendall trend analysis
install.packages("knitr")        # core R Markdown table printing (kable)
install.packages("kableExtra")   # publication-style HTML tables (multi-header, styling)
install.packages("DT")      # interactive HTML tables (search/sort/filter/pagination)
install.packages("htmltools") # HTML captions/labels (optional; helps with titles)
install.packages("patchwork")    # combining multiple ggplot2 plots into one layout
install.packages("httr2")        # making HTTP requests to web APIs
install.packages("jsonlite")     # parsing and generating JSON data
install.packages("tibble")       # modern and tidy data frames
install.packages("scales")   # Tools for labeling and formatting graphs
install.packages("mgcv")     # Functions for fitting Generalized Additive Models (GAMs)
```


```{r include=FALSE}
## Load packages

library(terra)          # raster processing
library(sf)             # spatial vector processing
library(dplyr)          # data manipulation
library(tidyr)          # data reshaping
library(exactextractr)  # raster to polygon extraction
library(units)          # unit consistency
library(readr)          # data import and export
library(ggplot2)        #visualizations 
library(viridis)        # smooth gradient collr for maps
library(tidyterra)      # ggplot2 geoms for terra objects
library(Kendall)        # trend related analysis tools
library(trend)         # trend related analysis tools
library(knitr)        # core R Markdown table printing (kable)
library(kableExtra)   # publication-style HTML tables (multi-header, styling)
library(DT)          # datatable() for interactive research tables in HTML
library(htmltools)   # tags$caption / HTML helpers for table captions 
library(patchwork)   #Patchwork operator  
library(httr2)       # making HTTP requests to web APIs
library(jsonlite)   # parsing and generating JSON data
library(tibble)     # modern and tidy data frames
library(scales)     # Tools for labeling and formatting graphs
library(mgcv)      # Functions for fitting Generalized Additive Models (GAMs)
```



```{r, include=FALSE}
# Working Directory Structure
getwd()
 # Set working directory
setwd("C:/Users/Mohamed Said/OneDrive/Desktop/Yale University/Capstone Project Directory")

# main project folders
dir.create("data", showWarnings = FALSE)
dir.create("archive", showWarnings = FALSE)
dir.create("clean_data", showWarnings = FALSE)
dir.create("gis_data", showWarnings = FALSE)
dir.create("metadata", showWarnings = FALSE)
dir.create("processed_data", showWarnings = FALSE)
dir.create("raw_data", showWarnings = FALSE)

# subfolders
dir.create(file.path("raw_data", "chirps"), recursive = TRUE, showWarnings = FALSE)
dir.create(file.path("raw_data", "faostat"), recursive = TRUE, showWarnings = FALSE)
dir.create(file.path("processed_data", "chirps"), recursive = TRUE, showWarnings = FALSE)
dir.create(file.path("processed_data", "gadm"), recursive = TRUE, showWarnings = FALSE)
dir.create(file.path("clean_data", "climate"), recursive = TRUE, showWarnings = FALSE)
dir.create(file.path("clean_data", "livestock"), recursive = TRUE, showWarnings = FALSE)
```


## Data Sources

This analysis integrates three primary datasets:

1. **CHIRPS v3 rainfall data (1981–2024)**  
   Satellite-gauge rainfall estimates from the Climate Hazards Center that provide 
   annual rainfall measurements across Africa at approximately 0.05° spatial resolution.

2. **FAOSTAT livestock stock data**  
   National livestock population estimates for Somalia including camels, cattle, 
   goats, and sheep. 
   FAOSTAT data were downloaded manually as a CSV from the FAO FAOSTAT portal (<https://www.fao.org/faostat/en/>) using the website’s Download/Export function after selecting Somalia, the relevant variables, and the study years.


3. **GADM administrative boundaries**  
   Vector shapefiles providing Somalia's national, regional, and district boundaries.

These datasets allow rainfall patterns to be summarized spatially and linked with 
national livestock population time series.

## Analytical Approach

The analysis follows five main steps:

1. **Rainfall preprocessing**
   CHIRPS annual rainfall rasters were downloaded for 1981–2024 and clipped to the 
   Somalia national boundary.

2. **Rainfall summaries**
   National and regional mean rainfall statistics were calculated for each year.

3. **Exploratory data analysis**
   Rainfall anomalies, distributions, and spatial variability were examined.

4. **Livestock time-series analysis**
   FAOSTAT livestock stocks were analyzed using growth rates, index values, and trend tests.

5. **Climate–livestock modeling**
   A generalized additive model (GAM) was used to test whether rainfall anomalies 
   (including lagged effects) are associated with livestock stock dynamics.
   
```{r, echo=TRUE}
# Download FAOSTAT data set (manual download):
# FAOSTAT (Food and Agriculture Organization of the United Nations)
# Portal: https://www.fao.org/faostat/en/
# I downloaded the dataset as a CSV from the FAOSTAT Data section,
# then saved the raw file locally (no API call).

faostat_raw <- read.csv("FAOSTAT_data_en_3-8-2026.csv")
# Save a copy into the project raw data folder
write.csv(faostat_raw, "raw_data/faostat/FAOSTAT_data_en_3-8-2026.csv", row.names = FALSE)


# Rename columns to consistent snake_case
names(faostat_raw) <- tolower(names(faostat_raw))
names(faostat_raw) <- gsub("\\.+", "_", names(faostat_raw))
names(faostat_raw) <- gsub("^_+|_+$", "", names(faostat_raw))  # trims leading/trailing _

somalia_livestock_stocks_1981_2024 <- faostat_raw %>%
  dplyr::rename(species = item)

# Sort by year and species
somalia_livestock_stocks_1981_2024 <- somalia_livestock_stocks_1981_2024[
  order(somalia_livestock_stocks_1981_2024$year,
        somalia_livestock_stocks_1981_2024$species),
]

csv_dest <- file.path("clean_data", "livestock", "somalia_livestock_stocks_1981_2024_faostat.csv")
write.csv(somalia_livestock_stocks_1981_2024, csv_dest, row.names = FALSE)

```



```{r download-GADM, echo=TRUE, eval=FALSE}
#Download Somali boundaries data set (vector data)
# URL to GADM Somalia level 0 shapefile (zip)
somalia_url <- "https://geodata.ucdavis.edu/gadm/gadm4.1/shp/gadm41_SOM_shp.zip"

# Local path for the zip file
zip_dest <- file.path("gis_data", "somalia_boundary.zip")

# Download the zip file
download.file(somalia_url, destfile = zip_dest, mode = "wb")

# Unzip the contents into gis_data folder
unzip(zip_dest, exdir = "gis_data")

# Clean up zip 
file.remove(zip_dest)

```


```{r}
#Load Somalia boundaries specially regional shape file or first level (vector data)
somalia_country <- sf::st_read("gis_data/gadm41_SOM_0.shp")
somalia_regions <- sf::st_read("gis_data/gadm41_SOM_1.shp")
somalia_districts <- sf::st_read("gis_data/gadm41_SOM_2.shp")
print(somalia_country)

names(somalia_districts)

```

## plot Somalia and the Administrative Regions of Somalia but
```{r}
# Plot the Somalia boundaries map
(Somali_map <- ggplot() +
  geom_sf(data = somalia_country, fill = "grey", color = "black", size = 0.5) +
  geom_sf(data = somalia_districts, color = "grey", size = 0.1, alpha = 0.5) +
  geom_sf(data = somalia_regions, aes(color = NAME_1), fill = NA, size = 0.6) +
  theme_minimal() +
  labs(title = " Administrative Regions and Districts of Somalia",
       subtitle = "Visualizing hierarchical administrative boundaries",
       caption = " Source: The Global Administrative Boundaries Database (GADM)", 
       color = "Regions")+
   theme(
    legend.position = "right",
    legend.key.size = unit(0.3, "cm"),
    plot.title = element_text(face = "bold", size = 14)
  ))

```


```{r download-chirps, echo=TRUE, eval=FALSE}
## download the chirps dataset

# CHIRPS v3 annual Africa data location
base_url <- "https://data.chc.ucsb.edu/products/CHIRPS/v3.0/annual/africa/tifs/"

years <- 1981:2024

# Download chirps into raw_data
for (y in years) {
  file_name <- paste0("chirps-v3.0.", y, ".tif")
  file_url <- paste0(base_url, file_name)
  file_dest <- file.path("raw_data","chirps", file_name)
  download.file(file_url, file_dest, mode = "wb")
}
```

Load the vector data files in the row folder using list.files and rename to Chirpsv3. Also work with new code to assign layers since the chirps data sets are stored in years from 1981 to 2024.

```{r}
# List all files
tif_files <- list.files(file.path("raw_data","chirps"), pattern="\\.tif$", full.names=TRUE)


# assignt to chirpsv3
chirpsv3 <- terra::rast(tif_files)

# Assign clean layer names (e.g., chirpsv3_1981, chirpsv3_1982, ...)
years <- 1981:2024
names(chirpsv3) <- paste0("chirpsv3_", years)

# Check the object
nlyr(chirpsv3)
res(chirpsv3)
names(chirpsv3)
```


```{r, include=FALSE}
print(paste("Raster CRS:", terra::crs(chirpsv3, describe = TRUE)$name))
print(paste("Vector CRS:", sf::st_crs(somalia_regions)$input))
```

# calucalte the summary statitics fo the data

```{r, include=FALSE}
summary(chirpsv3)
```

There is -9999 which is placeholder for the oceans pixels which has no CHIRPSV3 has not data. we will mask these values by replacing the -9999 to NA and dropping to select land bixelsn only

```{r, include=FALSE}
# Check value range
global(chirpsv3, range, na.rm = FALSE)
```

```{r, include=FALSE}
# Replace -9999 with NA
chirpsv3[chirpsv3 == -9999] <- NA
```

```{r, include=FALSE}
#Check value range again

global(chirpsv3, range, na.rm = TRUE)
```

```{r}
# Plot the masked raster using ggplot2 with tidyterra
ggplot() +
  geom_spatraster(data = chirpsv3) +
    ggplot2::scale_fill_gradientn(
    colors = c("#228A8D", "#56C666", "#DDE31C"), # low -> high
    na.value = "transparent",
    name = "Rainfall (mm)"
  ) +
  theme_void() +
  labs(title = "CHIRPS Annual Rainfall For Africa's Land Pixels Only")
```

```{r, include=FALSE}
# Ensure the CRS matches with the whole somalia and somalia regions 
somalia_country <- sf::st_transform(somalia_country, terra::crs(chirpsv3))
somalia_regions <- sf::st_transform(somalia_regions, terra::crs(chirpsv3))

somalia_country
somalia_regions
```

```{r, include=FALSE}
# CROP first (shrinks the rectangular box to Somalia's size)
chirps_som_crop <- terra::crop(chirpsv3, somalia_country)

# MASK second (uses the border to hide everything outside Somalia)
chirps_som_final <- terra::mask(chirps_som_crop, somalia_country)


# FINAL CLEANING
chirps_SOM_terrestrial <- chirps_som_final
chirps_SOM_terrestrial

#align regions to CRS to match the chirps_SOM_terrestrial 
somalia_regions <- st_transform(somalia_regions, crs(chirps_SOM_terrestrial))

# Save processed raster stack (Somalia only)
som_stack_path <- file.path("processed_data","chirps","chirps_SOM_terrestrial_1981_2024.tif")
terra::writeRaster(chirps_SOM_terrestrial, som_stack_path, overwrite = TRUE)
```

```{r}
# Plot the masked raster 
ggplot() +
  geom_spatraster(data = chirps_SOM_terrestrial) +
 scale_fill_gradientn(
    colours = c("#228A8D", "#56C666", "#DDE31C"), # low -> high
    na.value = "transparent",
    name = "Rainfall(mm)" ) +
  geom_sf(data = somalia_regions, fill = NA, linewidth = 0.4, color = "black") +
  geom_sf_text(data = somalia_regions, aes(label = NAME_1 ), size = 3, color = "darkblue") +
  theme_void() +
  labs(title = "CHIRPS Annual Rainfall Land Pixels Only for Somalia whole country ")  +
  theme(
    strip.text = element_text(face = "bold", size = 9),
    plot.title = element_text(face = "bold", size = 14),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 7)
  )

```

## Somalia National Mean Rainfall Time Series (1981–2024)

This gives a one value per year (mean rainfall over Somalia).

```{r}
# Somalia mean rainfall per year (over all pixels)
som_ts_mean <- terra::global(chirps_SOM_terrestrial, fun = "mean", na.rm = TRUE)
som_ts_mean <- tibble(
  year = 1981:2024,
  mean_rain = as.numeric(som_ts_mean[,1])
)

knitr::kable(
  som_ts_mean,
  format = "html",
  caption = "Somalia national mean annual rainfall (CHIRPS v3), 1981–2024"
) %>%
  kableExtra::kable_styling(
    full_width = T,
    bootstrap_options = c("striped", "condensed", "hover")
  )

#  time series Plot
ggplot(som_ts_mean, aes(year, mean_rain)) +
  geom_line(linewidth = 1, col = "darkblue") +
  theme_test() +
  labs(title = " Somali Annual Rainfall (CHIRPS v3) from 1981 to 2024",
       x = "Year", y = "Mean rainfall (mm)")  +
  theme(
    strip.text = element_text(face = "bold", size = 9),
    plot.title = element_text(face = "bold", size = 14),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 7)
  )
```

*Map of the Somalia Mean Annual Rainfall (CHIRPS v3, 1981–2024 MEAN RASTER)*

```{r}
# 1981–2024 MEAN RASTER (one layer)
chirps_SOM_mean_1981_2024 <- terra::app(chirps_SOM_terrestrial, fun = mean, na.rm = TRUE)

# Plot mean map + regions
ggplot() +
  geom_spatraster(data = chirps_SOM_mean_1981_2024) +
  scale_fill_gradientn(
    colours = c("#228A8D", "#56C666", "#DDE31C"),
    na.value = "transparent",
    name = "Mean Rainfall (mm)\n1981–2024"
  ) +
  geom_sf(data = somalia_regions, fill = NA, linewidth = 0.4, color = "black") +
  geom_sf_text(data = somalia_regions, aes(label = NAME_1), size = 3, color = "darkblue") +
  theme_void() +
  labs(title = "Somalia Mean Annual Rainfall (CHIRPS v3, 1981–2024)")  +
  theme(
    strip.text = element_text(face = "bold", size = 9),
    plot.title = element_text(face = "bold", size = 14),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 7)
  )
```

## Interannual Variability and Rainfall Anomalies

Using annual, country‐mean CHIRPS v3 rainfall for Somalia (1981–2024), we compute rainfall anomalies relative to the long-term national mean. The long-term baseline is calculated as the mean of the 44 annual national means, and each year’s anomaly is defined as:𝐴𝑡=𝑅𝑡−𝑅ˉ, where Rt​ where 𝑅𝑡 ​ is the annual national mean rainfall and𝑅 is the 1981–2024 mean. Positive anomalies indicate wetter-than-average years, while negative anomalies indicate drier-than-average years.

The anomaly series shows strong interannual variability with frequent switches between wet and dry years, indicating that Somalia’s national rainfall is dominated by year-to-year fluctuations. Extreme years stand out clearly: 1997 is the wettest outlier (anomaly ≈ +450.6 mm), while 2016 is among the driest years (anomaly ≈ −124.6 mm). Multi-year dry clusters are visible in the early 1990s (e.g., 1991–1993) and late 1990s–early 2000s (e.g., 1998–2001), while recent years remain highly variable, including strong wet anomalies in 2019 and 2023 and strong dry anomalies in 2021–2022. The anomaly (deviation from long-term mean) of Somalia since 1981

```{r}
som_ts_mean <- som_ts_mean %>%
  mutate(anomaly = mean_rain - mean(mean_rain, na.rm = TRUE))

datatable(som_ts_mean, rownames=FALSE, caption= "Somalia national mean annual rainfall (CHIRPS v3), 1981–2024")%>%
  DT::formatRound("mean_rain",4)

# lowest and highest rainfall years
extreme_rainfall_years <- som_ts_mean %>%
  filter(
    mean_rain == min(mean_rain, na.rm = TRUE) |
    mean_rain == max(mean_rain, na.rm = TRUE)
  ) %>%
  mutate(
    category = case_when(
      mean_rain == min(mean_rain, na.rm = TRUE) ~ "Lowest rainfall year",
      mean_rain == max(mean_rain, na.rm = TRUE) ~ "Highest rainfall year"
    )
  ) %>%
  select(category, year, mean_rain, anomaly)

extreme_rainfall_years %>%
  mutate(
    mean_rain = round(mean_rain, 4),
    anomaly = round(anomaly, 4)
  ) %>%
  knitr::kable(
    format = "html",
    rownames = FALSE,
    caption = "Somalia Lowest and Highest Annual Rainfall Years (CHIRPS v3), 1981–2024"
  ) %>%
  kableExtra::kable_styling(
    full_width = TRUE,
    bootstrap_options = c("striped", "condensed", "hover")
  )

ggplot(som_ts_mean, aes(year, anomaly)) +
  geom_hline(yintercept = 0) +
  geom_vline(
    data = extreme_rainfall_years,
    aes(xintercept = year, color = category),
    linetype = "dashed",
    linewidth = 0.8,
    show.legend = TRUE
  ) +
  geom_line(linewidth = 1, col = "#2C7FB8") +
  geom_text(
    data = extreme_rainfall_years,
    aes(
      x = year,
      y = max(som_ts_mean$anomaly, na.rm = TRUE) * 0.95,
      label = year,
      color = category
    ),
    size = 5,
    fontface = "bold",
    vjust = -0.2,
    show.legend = FALSE
  ) +
  scale_color_manual(values = c(
    "Lowest rainfall year" = "darkred",
    "Highest rainfall year" = "forestgreen"
  )) +
  theme_test() +
  labs(
    title = "Somalia Rainfall Anomalies vs 1981–2024 mean",
    x = "Year",
    y = "Anomaly (mm)",
    color = NULL
  ) +
  theme(
    strip.text = element_text(face = "bold", size = 9),
    plot.title = element_text(face = "bold", size = 14),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 7)
  )


# Save CLEAN climate table (national mean + anomaly)
write.csv(
  som_ts_mean,
  file.path("clean_data","climate","somalia_chirps_national_mean_anomaly_1981_2024.csv"),
  row.names = FALSE)

```

## Quick EDA summary (Somalia distribution across years)

The annual Somalia-wide mean rainfall series shows pronounced interannual variability, with clear separation between “wet” and “dry” years relative to the long-term mean. In the Wet/Dry classification, several early years were wetter than average (e.g., 1981–1982, 1987–1990), while prolonged dry conditions appear in the early 1990s (1991–1993) and again around the late 1990s to mid-2000s (many years from 1998–2005 classified as dry). Extreme years stand out clearly: 1997 is an exceptional wet outlier (mean ≈ 784 mm; anomaly ≈ +451 mm), whereas 2016 is among the driest years (mean ≈ 209 mm; anomaly ≈ −125 mm). Recent years show mixed conditions, including notable wet years (2019–2020 and 2023) and marked dry years (2021–2022).

**The spatial SD classification adds an important second dimension:** it quantifies whether rainfall in a given year was spatially uneven across Somalia. Years classified as high spatial variability (SD ≥ median) include several wetter-than-average years most notably 1997 (SD ≈ 476 mm), but also 1981, 1982, 1994, 2006, 2018, 2019, and 2023. This indicates that wet years often coincide with strong spatial contrasts, meaning that rainfall increases are not uniform across the country; instead, some areas receive substantially more rainfall than others. Conversely, many years classified as low spatial variability include several drought-like years (e.g., 1996, 2016, 2021–2022), suggesting that in dry years, conditions can be more uniformly dry across Somalia.

Overall, combining both classifications shows that Somalia’s rainfall variability is characterized not only by large year-to-year swings in the national mean, but also by systematic changes in within-country spatial heterogeneity: the wettest years tend to be associated with greater spatial disparity, while many of the driest years show lower spatial variability, consistent with widespread dryness

```{r}

# Summary per layer (year): mean, sd, min, max across pixels in Somalia
layer_stats <- terra::global(chirps_SOM_terrestrial,
                             fun = c("mean","sd","min","max"),
                             na.rm = TRUE)

layer_stats_df <- tibble(
  year = 1981:2024,
  mean = layer_stats$mean,
  sd   = layer_stats$sd,
  min  = layer_stats$min,
  max  = layer_stats$max
)

knitr::kable(
  layer_stats_df,
  format = "html",
  caption = "Somalia Interannual rainfall Statistics(1981–2024)"
) %>%
  kableExtra::kable_styling(
    full_width = TRUE,
    bootstrap_options = c("striped", "condensed", "hover")
  )

# Plot mean + variability
mean_plot<-ggplot(layer_stats_df, aes(year, mean)) +
  geom_line(linewidth = 1, col="#0B5D7A") +
  theme_test() +
  labs(title="Somalia Annual Mean Rainfall(1981–2024) ", y="Mean (mm)", x="Year")

sd_plot<-ggplot(layer_stats_df, aes(year, sd)) +
  geom_line(linewidth = 1, col="#D95F98") +
  theme_test() +
  labs(title="Somalia Spatial SD of Rainfall", y="SD (mm)", x="Year") 

mean_plot|sd_plot 
```

# Additional EDA: distribution + summary + extremes

These EDA results show that Somalia’s annual mean rainfall is highly variable across years, but most years cluster in a lower range, with one very extreme wet outlier.

The histogram suggests a right-skewed distribution: many years fall roughly between 250 and 380 mm, while a small number of wet years extend far above this range. The boxplot confirms this and shows that 1997 is a clear upper outlier, meaning it was unusually wet compared with the rest of the 1981–2024 record.

The summary statistics support this pattern. The long-term mean is 333.72 mm, while the median is lower at 302.44 mm, which is another sign of positive skewness caused by a few very wet years. The standard deviation (103.26 mm) is fairly large relative to the mean, indicating strong interannual variability in national rainfall.

The extremes table shows that the driest years were 2016, 2022, 1996, 2001, and 1999, all with strongly negative anomalies. The wettest years were 1997, 2023, 2019, 1982, and 1981, all with positive anomalies, with 1997 standing far above the others. In practical terms, this means Somalia’s rainfall regime is dominated less by a smooth long-term change and more by large year-to-year fluctuations, including occasional extreme wet years and several clustered dry years.

*conclution*: The exploratory analysis indicates that Somalia’s national annual rainfall distribution for 1981–2024 is positively skewed and characterized by substantial interannual variability. Most years cluster around moderate rainfall totals, but a few exceptionally wet years shift the upper tail of the distribution upward. The boxplot identifies 1997 as a strong wet outlier, while the extremes table shows that the driest years are concentrated mainly in the late 1990s, mid-2010s, and early 2020s. Overall, the rainfall series appears to be driven more by year-to-year variability and extreme events than by a stable central pattern.

```{r}
# 1) Distribution check: histogram
ggplot(som_ts_mean, aes(x = mean_rain)) +
  geom_histogram(bins = 15, fill = "forestgreen", color = "white", alpha = 0.8) +
  geom_vline(aes(xintercept = mean(mean_rain)), color = "darkred", linetype = "dashed", size = 1) +
  annotate("text", x = 450, y = 8, label = "Long-term Mean (333.72 mm)", color = "darkred") +
  scale_x_continuous(labels = unit_format(unit = "mm")) +
  labs(
    title = "Distribution of Somalia Annual Mean Rainfall",
    subtitle = "1981–2024 CHIRPS v3 Data showing right-skewed climate profile",
    x = "Annual Mean Rainfall (mm)",
    y = "Frequency (Number of Years)",
    caption = "Source: CHIRPS v3 Satellite-Gauge Estimates"
  ) +
  theme_test(base_size = 12)


# 2) Distribution check: boxplot
ggplot(som_ts_mean, aes(y = mean_rain)) +
  geom_boxplot(fill = "#99B9E9") +
  theme_test() +
  labs(
    title = "Boxplot of Somalia Annual Mean Rainfall (1981–2024)",
    x = NULL,
    y = "Mean rainfall (mm)"
  )

# 3) Summary table
rainfall_summary_table <- som_ts_mean %>%
  summarise(
    observations = n(),
    mean_rainfall_mm = round(mean(mean_rain, na.rm = TRUE), 2),
    sd_rainfall_mm = round(sd(mean_rain, na.rm = TRUE), 2),
    min_rainfall_mm = round(min(mean_rain, na.rm = TRUE), 2),
    median_rainfall_mm = round(median(mean_rain, na.rm = TRUE), 2),
    max_rainfall_mm = round(max(mean_rain, na.rm = TRUE), 2)
  )

knitr::kable(
  rainfall_summary_table,
  format = "html",
  caption = "Summary Statistics for Somalia National Annual Mean Rainfall (CHIRPS v3), 1981–2024"
) %>%
  kableExtra::kable_styling(
    full_width = TRUE,
    bootstrap_options = c("striped", "condensed", "hover")
  )

# 4) Extremes table
rainfall_extremes_table <- bind_rows(
  som_ts_mean %>%
    arrange(mean_rain) %>%
    slice(1:5) %>%
    mutate(category = "Lowest rainfall years"),
  som_ts_mean %>%
    arrange(desc(mean_rain)) %>%
    slice(1:5) %>%
    mutate(category = "Highest rainfall years")
) %>%
  select(category, year, mean_rain, anomaly)

knitr::kable(
  rainfall_extremes_table %>%
    mutate(
      mean_rain = round(mean_rain, 2),
      anomaly = round(anomaly, 2)
    ),
  format = "html",
  caption = "Lowest and Highest Rainfall Years in Somalia (CHIRPS v3), 1981–2024"
) %>%
  kableExtra::kable_styling(
    full_width = TRUE,
    bootstrap_options = c("striped", "condensed", "hover")
  )
```

## High vs Low Spatial Variability Classification (within-Somalia SD)

This step measures whether rainfall in a given year was spatially uniform or spatially heterogeneous across Somalia. For each year, you compute the spatial `SD` of rainfall across Somalia land pixels. You then classify the year as High spatial variability if `SD ≥` the median `SD` across 1981–2024, otherwise Low spatial variability. The Threshold_Definition column documents this rule.

*Results*: Years with the highest `SD` (e.g., 1997 at the top of your ranking) indicate rainfall was very uneven across Somalia—some areas received much more rainfall than others. Several other high-SD years (including some wet years) suggest that high national rainfall can coincide with strong geographic contrasts rather than uniform wetting. In contrast, `low-SD` years reflect more spatially consistent conditions (often widespread dryness or broadly similar rainfall patterns). This classification complements the Wet/Dry anomaly analysis by separating how wet/dry the year was nationally from how evenly rainfall was distributed spatially across the country

```{r}
sd_cut <- median(layer_stats_df$sd, na.rm = TRUE)

years_variability_table <- layer_stats_df %>%
  mutate(
   Spatial_Variability = ifelse(sd >= sd_cut, "High spatial variability", "Low spatial variability"),
    Threshold_Definition = ifelse(sd >= sd_cut, ">= median SD", "< median SD")
  ) %>%
  select(year, mean, sd, Spatial_Variability,Threshold_Definition ) %>%
  arrange(desc(sd)) %>%
  mutate(across(c(mean, sd), ~round(.x, 2)))

knitr::kable(
  years_variability_table,
  format = "html",
  caption = "Somalia Years Classified into Two Groups by Spatial Variability (SD), 1981–2024"
) %>%
  kableExtra::kable_styling(
    full_width = TRUE,
    bootstrap_options = c("striped", "condensed", "hover")
  )
```

## Wet vs Dry Year Classification (relative to 1981–2024 mean)

This step classifies each year as Wet or Dry based on whether Somalia’s national annual mean rainfall is above or below the 1981–2024 long-term mean. We first compute the anomaly `𝐴t= 𝑅𝑡−𝑅ˉA t ​=R t ​−`R ˉ , then assign Wet if `𝐴 𝑡 ≥ 0 A t ​≥0` and Dry if `𝐴 𝑡 < 0 A t ​ <0.` The extra column Threshold_Definition documents the rule used for each class.

*Results*: The classification shows frequent switching between Wet and Dry years, confirming strong interannual variability. Extreme conditions are clearly separated: 1997 is an exceptionally Wet year (very large positive anomaly), while 2016 is among the driest years (largest negative anomaly). Periods such as the early 1990s and late 1990s–early 2000s contain multiple Dry years, while recent years show sharp swings, including strongly Wet years (e.g., 2019, 2023) and strongly Dry years (e.g., 2021–2022). Overall, the Wet/Dry grouping provides an interpretable summary of national departures from the climatological baseline.

```{r}
years_class_table <- layer_stats_df %>%
  mutate(
    anomaly = mean - mean(mean, na.rm = TRUE),
    Rainfall_Class = ifelse(anomaly >= 0, "Wet", "Dry"),
    Threshold_Definition = ifelse(anomaly >= 0, ">= long-term mean", "< long-term mean")
  ) %>%
  select(year, mean, anomaly, Rainfall_Class, Threshold_Definition) %>%
  arrange(year) %>%
  mutate(across(c(mean, anomaly), ~round(.x, 2)))

knitr::kable(
  years_class_table,
  format = "html",
  caption = "Somalia Annual Rainfall Years Classified into Two Groups (Wet vs Dry), 1981–2024"
) %>%
  kableExtra::kable_styling(
    full_width = TRUE,
    bootstrap_options = c("striped", "condensed", "hover")
  )


ggplot(years_class_table, aes(x = factor(year), y = anomaly, fill = Rainfall_Class)) +
  geom_col() +
  scale_fill_manual(values = c("Wet" = "#2C7FB8", "Dry" = "#D95F0E")) +
  labs(
    title = "Annual Rainfall Anomalies in Somalia (1981–2024)",
    x = "Year",
    y = "Anomaly (mm)",
    fill = "Class"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)
  )
```
```{r}
## Wet vs Dry Year Classification (relative to 1981–2024 mean)

# 1. Create the classification table
years_class_table <- layer_stats_df %>%
  mutate(
    anomaly = mean - mean(mean, na.rm = TRUE),
    Rainfall_Class = ifelse(anomaly >= 0, "Wet", "Dry"),
    Threshold_Definition = ifelse(anomaly >= 0, ">= long-term mean", "< long-term mean")
  ) %>%
  select(year, mean, anomaly, Rainfall_Class, Threshold_Definition) %>%
  arrange(year) %>%
  mutate(across(c(mean, anomaly), ~round(.x, 2)))

# 2. Display the table
knitr::kable(
  years_class_table,
  format = "html",
  caption = "Somalia Annual Rainfall Years Classified into Two Groups (Wet vs Dry), 1981–2024"
) %>%
  kableExtra::kable_styling(
    full_width = TRUE,
    bootstrap_options = c("striped", "condensed", "hover")
  )

# 3. Plot with 5-year breaks
ggplot(years_class_table, aes(x = year, y = anomaly, fill = Rainfall_Class)) +
  geom_col() +
  scale_fill_manual(values = c("Wet" = "#2C7FB8", "Dry" = "#D95F0E")) +
  # Use scale_x_continuous instead of factor(year) to allow for numeric breaks
  scale_x_continuous(breaks = seq(1981, 2024, by = 5)) +
  labs(
    title = "Annual Rainfall Anomalies in Somalia (1981–2024)",
    subtitle = "Highlighting 5-year intervals for improved readability",
    x = "Year",
    y = "Anomaly (mm)",
    fill = "Class"
  ) +
  theme_pubr() + 
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold", size = 14),
    axis.title = element_text(face = "bold"),
    panel.grid.major.y = element_line(color = "grey90", linetype = "dashed") # subtle horizontal guides
  )
```

## OLS Linear Trend Analysis of National Rainfall

Using annual, country‐mean CHIRPS v3 rainfall for Somalia (1981–2024), we fit an ordinary least squares model: *mean_raint​=β0​+β1​t+εt*​.

The estimated linear trend is β₁ = −0.0768 mm·yr⁻¹ (SE = 1.240), indicating an extremely small decrease. Scaled over the 44-year record, this corresponds to an expected change of approximately −3.4 mm, which is negligible relative to the observed interannual variability (residual SD ≈ 104.5 mm).

Statistical evidence for a monotonic trend is absent: t(42) = −0.062, p = 0.951 for β₁. The model explains essentially none of the variance in annual mean rainfall (R² = 9.1×10⁻⁵; adjusted R² = −0.0237), consistent with rainfall variability being dominated by year-to-year fluctuations rather than a detectable linear change both increase or decrease over the study period in the CHIRPS series.

```{r}
fit <- lm(mean_rain ~ year, data = som_ts_mean)
summary(fit)
  
ggplot(som_ts_mean, aes(year, mean_rain)) +
  geom_line(linewidth = 1, col="#0B5D7A") +
  geom_smooth(method = "lm", se = TRUE) +
  theme_test() +
  labs(title="Somalia Rainfall Trend (linear fit)", x="Year", y="Mean (mm)")

```

## National Monotonic Trend in Mean Rainfall (1981–2024; MK + Sen)

This section tests whether Somalia’s national annual mean rainfall exhibits a monotonic long-term change over 1981–2024. We use the Mann–Kendall rank test to evaluate trend significance and Sen’s slope to estimate a robust trend magnitude. These non-parametric methods are widely used in hydro-climate trend detection because they do not require normality and are less sensitive to extreme wet/dry years than OLS-based inference.

*Results*: At the national scale, there is no evidence of a monotonic rainfall trend over 1981–2024 (Mann–Kendall τ ≈ 0.002, p ≈ 0.992). Sen’s slope is near zero (≈ 0.020 mm·yr⁻¹; 95% CI ≈ −1.983 to 1.943 mm·yr⁻¹), indicating negligible long-term change relative to interannual variability.

```{r}
## National rainfall trend (Somalia, CHIRPS v3, 1981–2024): Mann–Kendall + Sen slope

# Mann–Kendall trend test (non-parametric)
mk_nat  <- Kendall::MannKendall(som_ts_mean$mean_rain)

# Sen slope (robust slope estimate + 95% CI)
sen_nat <- trend::sens.slope(som_ts_mean$mean_rain)

cat("Mann–Kendall τ statistic:", mk_nat$tau, " | p-value:", mk_nat$sl, "\n")

cat("Sen’s slope estimate (mm/year):", sen_nat$estimates,
    " | 95% CI bounds:", sen_nat$conf.int, "\n")

```

## Regional Annual Mean Rainfall Time Series (1981-2024)

```{r}
## Regional mean rainfall  time series Plot (18 regions, 1981–2024)
region_ts_mean <- exactextractr::exact_extract(
  chirps_SOM_terrestrial, somalia_regions, "mean")
region_ts_mean <- tibble(
  Region = somalia_regions$NAME_1,
  region_ts_mean)
names(region_ts_mean)[-1] <- 1981:2024
region_ts_mean <- region_ts_mean %>%
  pivot_longer(
    cols = -Region,
    names_to = "year",
    values_to = "mean_rain"
  ) %>%
  mutate(year = as.integer(year))

datatable(
  region_ts_mean,
  rownames = FALSE,
  caption = "Somalia regional mean annual rainfall (CHIRPS v3), 1981–2024)"
) %>%
  formatRound("mean_rain", 2)

# Plot the 18 region mean rainfall

ggplot(region_ts_mean, aes(year, mean_rain)) +
  geom_line(linewidth = 1, col = "steelblue") +
  facet_wrap(~Region, scales = "free_y", ncol = 3) +
  theme_minimal() +
  labs(
    title = "Regional Annual Rainfall in Somalia (CHIRPS v3) from 1981 to 2024",
    x = "Year",
    y = "Mean rainfall (mm)"
  ) +
  theme(
    strip.text = element_text(face = "bold", size = 9),
    plot.title = element_text(face = "bold", size = 14),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 7)
  )


write.csv(region_ts_mean,
          file.path("clean_data","climate","somalia_chirps_regional_mean_1981_2024.csv"),
          row.names = FALSE)
```

# Regional Rainfall Statistics (1981–2024)

```{r}
# Summary per region across years: mean, sd, min, max
region_stats <- region_ts_mean %>%
  group_by(Region) %>%
  summarise(
    mean = mean(mean_rain, na.rm = TRUE),
    sd   = sd(mean_rain, na.rm = TRUE),
    min  = min(mean_rain, na.rm = TRUE),
    max  = max(mean_rain, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(mean)) %>%
  mutate(across(c(mean, sd, min, max), round, 2))

knitr::kable(
  region_stats,
  format = "html",
  caption = "Regional Rainfall Statistics in Somalia (1981–2024)"
) %>%
  kableExtra::kable_styling(
    full_width = TRUE,
    bootstrap_options = c("striped", "condensed", "hover")
  )

# Plot mean + variability across regions
mean_plot <- ggplot(region_stats, aes(reorder(Region, mean), mean)) +
  geom_col(fill = "#0B5D7A") +
  coord_flip() +
  theme_test() +
  labs(
    title = "Mean Somali Regional Rainfall (1981–2024)",
    x = NULL, y = "Mean (mm)"
  )

sd_plot <- ggplot(region_stats, aes(reorder(Region, sd), sd)) +
  geom_col(fill = "#A63D40") +
  coord_flip() +
  theme_test() +
  labs(
    title = " Somali Regional Rainfall Variability (1981–2024)",
    x = NULL, y = "SD (mm)"
  )

mean_plot | sd_plot
```

## Regional Rainfall Anomalies in Somalia Heatmap (1981–2024)

```{r}
region_ts_mean <- region_ts_mean %>%
  group_by(Region) %>%
  mutate(
    anomaly = mean_rain - mean(mean_rain, na.rm = TRUE)
  ) %>%
  ungroup()

ggplot(region_ts_mean, aes(year, Region, fill = anomaly)) +
  geom_tile(color = "white", linewidth = 0.15) +
  scale_fill_gradient2(
    low = "#A63D40",
    mid = "#F7F7F7",
    high = "#2C6E91",
    midpoint = 0,
    name = "Anomaly (mm)"
  ) +
  theme_minimal() +
  labs(
    title = "Regional Rainfall Anomalies in Somalia (1981–2024)",
    x = "Year",
    y = NULL
  ) +
  theme(
    axis.text.y = element_text(size = 9),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
    plot.title = element_text(face = "bold", size = 13),
    panel.grid = element_blank()
  )
```

## Regional Spatial Distribution and classification of Long-Term Mean Rainfall (1981–2024 Map)

This section summarizes long-term spatial patterns in Somalia’s rainfall. First, we collapse the annual CHIRPS layers (1981–2024) into a single raster representing the multi-year mean rainfall. Next, we compute the mean rainfall for each administrative region using zonal extraction. Finally, we identify the wettest and driest regions (highest vs lowest mean) and visualize them on the mean rainfall map, keeping all other regions in grey for context.

*Results:* The 1981–2024 CHIRPS climatology reveals strong spatial heterogeneity in Somalia’s rainfall. The highest long-term regional means occur in the southern regions specillay (Jubbada Hoose, Jubbada Dhexe, Bay, and Shabeellaha Hoose; \~557–592 mm), while the lowest means occur in the northeast and east (Bari \~125 mm; Sool, Nugaal, Sanaag \~190–199 mm).

Overall, the wettest region receives roughly 4–5 times the rainfall of the driest region, indicating a pronounced spatial rainfall gradient across the country. Based on 1981–2024 regional mean rainfall, Somalia’s regions can be grouped into very wet (≥500 mm), moderately wet (400–499 mm), semi-arid (250–399 mm), and arid (\<250 mm) zones. The specific regions falling within each category are reported in Table *[X]*.

```{r}
# Long-term mean rainfall by region from the regional time series
regions_mean_table <- region_ts_mean %>%
  group_by(Region) %>%
  summarise(
    Mean_Rainfall_mm_1981_2024 = round(mean(mean_rain, na.rm = TRUE), 2),
    SD_Rainfall = round(sd(mean_rain, na.rm = TRUE), 2),
    .groups = "drop"
  ) %>%
  mutate(
    Rainfall_Class = case_when(
      Mean_Rainfall_mm_1981_2024 >= 500 ~ "Very wet",
      Mean_Rainfall_mm_1981_2024 >= 400 ~ "Moderately wet",
      Mean_Rainfall_mm_1981_2024 >= 250 ~ "Semi-arid",
      TRUE                              ~ "Arid"
    ),
    Threshold_Definition = case_when(
      Mean_Rainfall_mm_1981_2024 >= 500 ~ ">= 500 mm",
      Mean_Rainfall_mm_1981_2024 >= 400 ~ "400–499 mm",
      Mean_Rainfall_mm_1981_2024 >= 250 ~ "250–399 mm",
      TRUE                              ~ "< 250 mm"
    )
  ) %>%
  arrange(desc(Mean_Rainfall_mm_1981_2024))

knitr::kable(
  regions_mean_table,
  format = "html",
  caption = "Somalia Regions Classified by Mean Annual Rainfall (CHIRPS v3, 1981–2024)"
) %>%
  kableExtra::kable_styling(
    full_width = TRUE,
    bootstrap_options = c("striped", "condensed", "hover")
  )

```

## Heat Map for the Regional Rainfall classification in Somalia (1981-2024)

```{r}
ggplot(
  regions_mean_table,
  aes(x = "1981–2024", y = reorder(Region, Mean_Rainfall_mm_1981_2024), fill = Mean_Rainfall_mm_1981_2024)
) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = Rainfall_Class), size = 3.2, color = "black", fontface = "bold") +
  scale_fill_gradientn(
    colours = c("#8C510A", "#D8B365", "#F6E8C3", "#5AB4AC", "#01665E"),
    name = "Mean rainfall (mm)"
  ) +
  theme_minimal() +
  labs(
    title = "Regional Rainfall Classification in Somalia (1981–2024)",
    x = NULL,
    y = NULL
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.text.x = element_text(face = "bold"),
    axis.text.y = element_text(size = 9),
    panel.grid = element_blank()
  )
```

### Regional Differences: Wettest vs Driest Regions

Zonal means of the 1981–2024 CHIRPS climatology indicate that Jubbada Hoose is the wettest administrative region (mean ≈ 592 mm/year), whereas Bari is the driest (mean ≈ 125 mm/year). The wettest region therefore receives roughly 4.7 times the average rainfall of the driest region, highlighting pronounced spatial heterogeneity in Somalia’s long-term rainfall distribution. In the map, the mean rainfall surface is shown with the wettest and driest regions emphasized (red and blue outlines, respectively), while other regions are retained in grey to preserve geographic context. Given that *Jubbada Hoose* lies in southern Somalia and *Bari* lies in the northeast/east, these results are consistent with a strong spatial rainfall gradient across the country, with wetter conditions in the south and markedly drier conditions toward the northeast/east

```{r}
# Identify wettest and driest regions from regions_mean_table
hi_name <- regions_mean_table$Region[which.max(regions_mean_table$Mean_Rainfall_mm_1981_2024)]
lo_name <- regions_mean_table$Region[which.min(regions_mean_table$Mean_Rainfall_mm_1981_2024)]

somalia_regions$hi_lo <- ifelse(
  somalia_regions$NAME_1 == hi_name, "Highest mean rainfall",
  ifelse(somalia_regions$NAME_1 == lo_name, "Lowest mean rainfall", "Other")
)

somalia_regions %>%
  st_drop_geometry() %>%
  filter(hi_lo != "Other") %>%
  select(Region = NAME_1, hi_lo) %>%
  left_join(
    regions_mean_table %>%
      select(Region, Mean_Rainfall_mm_1981_2024),
    by = "Region"
  ) %>%
  knitr::kable(
    format = "html",
    caption = "Wettest and driest regions in Somalia based on mean annual rainfall (1981–2024)"
  ) %>%
  kableExtra::kable_styling(
    full_width = FALSE,
    bootstrap_options = c("striped", "condensed", "hover")
  )

# Plot: mean raster + grey regions + highlight highest/lowest
ggplot() +
  geom_spatraster(data = chirps_SOM_mean_1981_2024) +
  scale_fill_gradientn(
    colours = c("#228A8D", "#56C666", "#DDE31C"),
    na.value = "transparent",
    name = "Mean Rainfall (mm)\n1981–2024"
  ) +
  geom_sf(data = somalia_regions, fill = NA, color = "grey60", linewidth = 0.35) +
  geom_sf(
    data = subset(somalia_regions, hi_lo != "Other"),
    aes(color = hi_lo),
    fill = NA,
    linewidth = 1.2
  ) +
  scale_color_manual(
    values = c(
      "Highest mean rainfall" = "#A63D90",
      "Lowest mean rainfall" = "#2C6E99"
    ),
    name = "Region class"
  ) +
  geom_sf_text(
    data = subset(somalia_regions, hi_lo != "Other"),
    aes(label = NAME_1, color = hi_lo),
    size = 3
  ) +
  theme_void() +
  labs(
    title = "Somalia Mean Rainfall (1981–2024): Wettest vs Driest Region"
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    legend.position = "right"
  )
```

# Regional Trend Analysis: MK + Sen + OLS

```{r}
## Regional trend analysis: MK + Sen + OLS

region_trend <- region_ts_mean %>%
  group_by(Region) %>%
  summarise(
    mk_tau    = Kendall::MannKendall(mean_rain)$tau,
    mk_p      = Kendall::MannKendall(mean_rain)$sl,
    sen_slope = trend::sens.slope(mean_rain)$estimates,
    ols_slope = coef(lm(mean_rain ~ year))[2],
    ols_p     = summary(lm(mean_rain ~ year))$coefficients[2,4],
    .groups = "drop"
  ) %>%
  mutate(
    Trend_Class = case_when(
      mk_p < 0.05 & sen_slope > 0 ~ "Significant increase",
      mk_p < 0.05 & sen_slope < 0 ~ "Significant decrease",
      TRUE                        ~ "No significant trend"
    )
  ) %>%
  mutate(across(c(mk_tau, mk_p, sen_slope, ols_slope, ols_p), ~round(.x, 4))) %>%
  arrange(desc(sen_slope))

knitr::kable(
  region_trend,
  format = "html",
  caption = "Regional rainfall trend results in Somalia (1981–2024)"
) %>%
  kableExtra::kable_styling(
    full_width = TRUE,
    bootstrap_options = c("striped", "condensed", "hover")
  )

```

# Regional Rainfall Trend CLassificaiton in Somalia

```{r}
somalia_regions <- left_join(
  somalia_regions,
  region_trend,
  by = c("NAME_1" = "Region")
)

ggplot(somalia_regions) +
  geom_sf(aes(fill = Trend_Class), color = "white", linewidth = 0.35) +
  geom_sf_text(aes(label = NAME_1), size = 2.8, color = "black") +
  scale_fill_manual(
    values = c(
      "Significant increase" = "#2C6E91",
      "Significant decrease" = "#A63D40",
      "No significant trend" = "grey75"
    ),
    name = "Trend class"
  ) +
  theme_void() +
  labs(title = "Regional Rainfall Trend Classification in Somalia (1981–2024)") +
  theme(
    plot.title = element_text(face = "bold", size = 13),
    legend.position = "right"
  )
```

# FOASTAT DATA ANALYSIS

## Foastat data Cleaning

```{r}
# Quick data checks 
names(somalia_livestock_stocks_1981_2024)
str(somalia_livestock_stocks_1981_2024)
summary(somalia_livestock_stocks_1981_2024)


# converting the the values as.numeric and year as.integer
somalia_livestock_stocks_1981_2024 <- somalia_livestock_stocks_1981_2024 %>%
  mutate(
    year  = as.integer(year),
    value = as.numeric(value),
    species = as.factor(species),
    species = as.factor(species) 
  )

# Check duplicates by species-year (should usually be 1 per species per year)
dup_check <- somalia_livestock_stocks_1981_2024 %>%
  count(species, year) %>%
  filter(n > 1)

# Missing values overview
colSums(is.na(somalia_livestock_stocks_1981_2024))

# Flag distribution (data quality context)
somalia_livestock_stocks_1981_2024 %>%
  count(flag, flag_description, sort = TRUE)
```

## Time series indicators

To characterize livestock stock dynamics over time (trend, volatility, and growth), we compute standard year-to-year change metrics within each species (item). After sorting by year, we derive: (i) absolute YoY change *`ΔVt​=Vt​−Vt−1`​* (ii) YoY percent change *`gt​=(Vt​/Vt−1​−1)×100`* and (iii) a baseline index (1981=100) *`It​=(Vt​/V1981​)×100`*. These indicators enable cross-species comparison on both absolute and relative scales while preserving the original measurement unit (headcount).

**Results***:*

-   **yoy_change (annual absolute change, heads)**: The largest absolute year-to-year declines are concentrated in the early 1990s and are strongest in small ruminants. Goats show the biggest shocks: in 1991 yoy_change = −4,500,000 (stocks drop from 18,500,000 to 14,000,000), followed by 1992 yoy_change = −4,000,000 (down to 10,000,000).

-   Sheep also decline sharply in 1991 with yoy_change = −2,000,000 (13,000,000 → 11,000,000). The contraction is also visible in other species but at smaller magnitudes, including cattle in 1990 (yoy_change = −800,000, 4,800,000 → 4,000,000) and camels in 1991 (yoy_change = −500,000, 6,700,000 → 6,200,000).

-   Immediately after the contraction, the series shows clear rebound jumps: cattle (1993) +900,000 (3,300,000 → 4,200,000), sheep (1993) +1,500,000 (10,000,000 → 11,500,000), and goats (1993) +1,000,000 (10,000,000 → 11,000,000). Overall, the time series is dominated by a sharp contraction (1990–1992) followed by a rebound (1993–1994), with the largest absolute shocks occurring in goats and sheep.

-   **`yoy_pct` (Annual Growth/Decline Rate, %):** The strongest year-to-year percentage contractions are tightly concentrated in **1990–1992**, and they are most severe for **goats and sheep**. Goats experience the sharpest drops in **1991** and **1992**, indicating an unusually large contraction over two consecutive years. Sheep also show their steepest negative percentage change in **1991**, with continued decline into **1992**. Cattle show their largest percentage decline in **1990**, followed by another substantial negative rate in **1991**. Camels decline as well in **1991–1992**, but the percentage losses are clearly smaller than for the other species.

-   The strongest positive percentage growth rates occur immediately after this contraction, with the most pronounced rebound in **1993–1994**. Cattle show the most intense recovery rate in **1993**, consistent with a rebound from a low point reached in 1992. Sheep show strong positive growth in **1993** and again in **1994**, indicating a two-year recovery phase. Goats also show a marked rebound in **1993** and **1994**, reversing part of the early-1990s decline. In contrast, camels exhibit only moderate positive growth rates across the rebound period, suggesting comparatively smoother adjustment.

-   Overall, the `yoy_pct` series is dominated by a **high-volatility episode around the early 1990s**—a rapid contraction followed by a rapid recovery—while later decades show **much smaller year-to-year percentage movements**, even when livestock levels remain large

-   **Index_1981 — long-run position relative to the 1981 baseline:** For `index_1981`, your results summarize the **long-run position of each herd relative to its own 1981 baseline**, so it’s the cleanest way to describe “where the series ends up” after all short-run volatility is absorbed. By 2024, **camels** sit at `index_1981 = 126.25833`, meaning the camel stock level is about **26% higher than in 1981**, while **cattle** end at `107.04699`, indicating a **modest net gain** of roughly **7% above** the baseline. In contrast, the small-ruminant series remain structurally below their starting point: **goats** end at `62.78391` (roughly **37% below** the 1981 level) and **sheep** at `73.96233` (about **26% below** baseline). In results terms, the index makes the long-run divergence explicit—**camels and cattle show net expansion over 1981–2024**, while **goats and sheep do not recover to baseline by 2024**, even though they exhibit periods of short-run rebound in the intermediate years.


```{r}
# Load necessary library for labeling
# install.packages("ggrepel")
library(ggrepel)

# Prepare data for end-of-line labels
label_data <- ts_tbl %>%
  group_by(species) %>%
  filter(year == max(year))

# 1) Plot: Levels trend (Total Stocks)
ggplot(ts_tbl, aes(x = year, y = value, color = species)) +
  geom_line(linewidth = 1.2) +
  # Adding direct labels at the end of the lines
  geom_text_repel(data = label_data, aes(label = species), 
                  nudge_x = 2, direction = "y", segment.color = 'grey50') +
  scale_y_continuous(labels = label_comma()) +
  scale_x_continuous(breaks = seq(1981, 2024, by = 5), limits = c(1981, 2028)) +
  scale_color_brewer(palette = "Set1") + # High contrast palette
  labs(
    title = "Livestock Population Trends in Somalia",
    subtitle = "Annual headcount (1981-2024)",
    x = "Year",
    y = "Total Stock (Number of Heads)",
    caption = "Source: FAOSTAT | Yale Environmental Data Science"
  ) +
  theme_pubr() +
  theme(legend.position = "none") # Legend removed because we used direct labels

# 2) Plot: Growth rate (YoY %)
ggplot(ts_tbl, aes(x = year, y = yoy_pct, color = species)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  geom_line(linewidth = 1, alpha = 0.8) +
  scale_x_continuous(breaks = seq(1981, 2024, by = 5)) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "Year-over-Year Growth Rate by Species",
    subtitle = "Highlighting high-volatility episode in the early 1990s",
    x = "Year", y = "YoY Growth (%)"
  ) +
  theme_pubr() +
  theme(legend.position = "bottom")

# 3) Plot: Indexed livestock stocks (1981 = 100)
ggplot(ts_tbl, aes(x = year, y = index_1981, color = species)) +
  geom_hline(yintercept = 100, linetype = "dotted", color = "grey40") +
  geom_line(linewidth = 1.2) +
  # Direct labels again for clarity in the index chart
  geom_text_repel(data = label_data, aes(label = species), 
                  nudge_x = 2, direction = "y") +
  scale_x_continuous(breaks = seq(1981, 2024, by = 5), limits = c(1981, 2028)) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "Indexed Livestock Stocks (Base Year 1981 = 100)",
    subtitle = "Long-term divergence: Camels/Cattle expansion vs. Small Ruminant decline",
    x = "Year", y = "Index Value"
  ) +
  theme_pubr() +
  theme(legend.position = "none")
```


```{r}
ts_tbl <- somalia_livestock_stocks_1981_2024 %>%
  arrange(species, year) %>%
  group_by(species) %>%
  mutate(
    yoy_change = value - lag(value),
    yoy_pct    = 100 * (value / lag(value) - 1),
    index_1981 = 100 * (value / first(value))
  ) %>%
  ungroup()

# Interactive table for the time-series indicators
DT::datatable(ts_tbl, rownames = FALSE, caption  = "Somalia livestock stocks (FAOSTAT): time-series indicators by species, 1981–2024",
  options  = list(pageLength = 25, scrollX = TRUE)
)
# 1) Plot: levels trend time (stocks)
ggplot(ts_tbl, aes(x = year, y = value, color = species)) +
  geom_line(linewidth = 1) +
  scale_y_continuous(labels = label_comma()) +
  scale_color_viridis_d(option = "plasma", name = "Livestock Species") +
  labs(
    title = "Livestock Population Trends in Somalia",
    subtitle = "Comparative analysis of Cattle, Goats, Camels, and Sheep (1981-2024)",
    x = "Year",
    y = "Total Stock (Number of Heads)",
    caption = "Source: FAOSTAT Data processed in EDS Capstone Script"
  ) +
  theme_test() +
  theme(legend.position = "bottom")


# 2) Plot: growth rate (YoY %)
ggplot(ts_tbl, aes(x = year, y = yoy_pct, color= species)) +
  geom_hline(yintercept = 0) +
  geom_line(linewidth = 1) +
  theme_test()+
  scale_y_continuous(labels = label_comma()) +
  labs(
    title = "Year-over-year growth rate (FAOSTAT), 1981–2024",
    x = "Year", y = "YoY %"
  )

# 3) Plot: index (1981 = 100) for comparison across animals
ggplot(ts_tbl, aes(x = year, y = index_1981, color = species)) +
  geom_line(linewidth = 1) +
theme_test()+
  labs(
    title = "Indexed livestock stocks (1981 = 100)",
    x = "Year", y = "Index"
  )
```

# Exploratory summary of livestock stock distributions and trends

The livestock EDA shows that the four species differ clearly in both stock size and variability over 1981–2024. Goats have the highest average stock level (13.63 million) and also the largest variability (SD = 3.11 million; CV = 22.83%), indicating that goat stocks changed more strongly over time than the other species. Sheep also have relatively high stock levels (mean = 12.20 million) with moderate variability (CV = 12.72%). In contrast, camels and cattle are more stable, with lower coefficients of variation (7.58% and 9.77%, respectively), suggesting less fluctuation across the study period.

The histograms and boxplots support these results. Goat stocks show the widest spread and the clearest upper values, while sheep also display broad dispersion. Camel stocks are more tightly clustered, consistent with a relatively steady long-term increase. Cattle show a narrower distribution centered around about 4.8–5.0 million, though with a few lower years.

The extremes table highlights important turning points. 1992 is the lowest stock year for camels, cattle, and goats, suggesting a common livestock shock or contraction during the early 1990s. For sheep, the minimum occurs in 2024, while the highest value occurs in 2005. Goat stocks reached their maximum much earlier, in 1988, and then declined substantially afterward. Camels, by contrast, reached their maximum in 2024, which is consistent with sustained long-run growth.

The trend table confirms these differences statistically. Camels show a clear and significant long-term increase (Sen’s slope ≈ +35,108 head/year; MK p \< 0.001). Goats show a strong and significant long-term decrease (Sen’s slope ≈ −92,597 head/year; MK p \< 0.001). Cattle show no statistically significant trend, indicating relative long-run stability. Sheep show a negative OLS slope, but the Mann–Kendall result is not significant at the 5% level, so this decline should be interpreted cautiously. Overall, the EDA suggests a long-run divergence in Somalia’s livestock structure: camels increased, goats declined, cattle remained broadly stable, and sheep showed a weaker downward tendency.

```{r}

# 1) Distribution check: histogram by species
ggplot(somalia_livestock_stocks_1981_2024, aes(x = value)) +
  geom_histogram(bins = 12, fill = "steelblue", color = "white") +
  facet_wrap(~ species, scales = "free") +
  theme_test() +
  labs(
    title = "Distribution of Livestock Stocks by Species in Somalia (FAOSTAT), 1981–2024",
    x = "Stock level",
    y = "Count"
  )

# 2) Distribution check: boxplot by species
ggplot(somalia_livestock_stocks_1981_2024, aes(x = species, y = value, fill = species)) +
  geom_boxplot() +
  theme_test() +
  labs(
    title = "Boxplot of Livestock Stocks by Species in Somalia (FAOSTAT), 1981–2024",
    x = NULL,
    y = "Stock level"
  ) +
  theme(legend.position = "none")

# 3) Summary table by species
livestock_summary_table <- somalia_livestock_stocks_1981_2024 %>%
  group_by(species) %>%
  summarise(
    observations = n(),
    mean_stock = round(mean(value, na.rm = TRUE), 2),
    sd_stock = round(sd(value, na.rm = TRUE), 2),
    min_stock = round(min(value, na.rm = TRUE), 2),
    median_stock = round(median(value, na.rm = TRUE), 2),
    max_stock = round(max(value, na.rm = TRUE), 2),
    cv_percent = round(100 * sd(value, na.rm = TRUE) / mean(value, na.rm = TRUE), 2),
    .groups = "drop"
  ) %>%
  arrange(desc(mean_stock))

knitr::kable(
  livestock_summary_table,
  format = "html",
  caption = "Summary Statistics for Somalia Livestock Stocks by Species (FAOSTAT), 1981–2024"
) %>%
  kableExtra::kable_styling(
    full_width = TRUE,
    bootstrap_options = c("striped", "condensed", "hover")
  )

# 4) Extremes table by species
livestock_extremes_table <- somalia_livestock_stocks_1981_2024 %>%
  group_by(species) %>%
  summarise(
    lowest_year = year[which.min(value)],
    lowest_stock = round(min(value, na.rm = TRUE), 2),
    highest_year = year[which.max(value)],
    highest_stock = round(max(value, na.rm = TRUE), 2),
    .groups = "drop"
  )

knitr::kable(
  livestock_extremes_table,
  format = "html",
  caption = "Lowest and Highest Livestock Stock Years by Species in Somalia (FAOSTAT), 1981–2024"
) %>%
  kableExtra::kable_styling(
    full_width = TRUE,
    bootstrap_options = c("striped", "condensed", "hover")
  )

# 5) Species trend table
livestock_species_trend_table <- somalia_livestock_stocks_1981_2024 %>%
  group_by(species) %>%
  summarise(
    ols_slope = coef(lm(value ~ year))[2],
    ols_p = summary(lm(value ~ year))$coefficients[2, 4],
    mk_tau = Kendall::MannKendall(value)$tau,
    mk_p = Kendall::MannKendall(value)$sl,
    sen_slope = trend::sens.slope(value)$estimates,
    .groups = "drop"
  ) %>%
  mutate(
    Trend_Class = case_when(
      mk_p < 0.05 & sen_slope > 0 ~ "Significant increase",
      mk_p < 0.05 & sen_slope < 0 ~ "Significant decrease",
      TRUE ~ "No significant trend"
    ),
    across(c(ols_slope, ols_p, mk_tau, mk_p, sen_slope), ~ round(.x, 4))
  ) %>%
  arrange(desc(sen_slope))

knitr::kable(
  livestock_species_trend_table,
  format = "html",
  caption = "Trend Results for Somalia Livestock Stocks by Species (FAOSTAT), 1981–2024"
) %>%
  kableExtra::kable_styling(
    full_width = TRUE,
    bootstrap_options = c("striped", "condensed", "hover")
  )
```


**Model**: Livestock–climate linkages were evaluated using a generalized additive model (GAM) to allow non-linear relationships and lagged rainfall effects. The response variable was log-transformed FAOSTAT national livestock stocks (camels, cattle, goats, sheep). Predictors included (i) a species-specific smooth of time to capture long-run, non-linear trajectories and (ii) species-specific smooth terms for national CHIRPS rainfall anomalies at 0-, 1-, and 2-year lags. Models were estimated in mgcv using REML for smoothing-parameter selection. This approach is designed to detect “curved” (non-linear) responses and delayed effects rather than forcing a straight-line relationshi.

**Method** : Annual CHIRPS national rainfall anomalies were used to construct lagged anomaly terms (t, t−1, t−2), which were then merged by year to the FAOSTAT livestock series. Early years without complete lag structure were excluded, yielding n = 168 observations. Stocks were modeled on the log scale to stabilize variance and support multiplicative interpretation. Model adequacy was assessed with standard GAM diagnostics (residual plots, fitted vs observed, and basis-dimension checks). Because the rainfall series is national (not regional), the model tests whether national rainfall departures from average are associated with national livestock dynamics.

**Results and interpretation**: The GAM explains most of the variation in livestock stocks (adjusted R² = 0.979; deviance explained = 98.4%), indicating that the model reproduces the observed national livestock trajectories closely. Species-level differences are strong (all p \< 2×10⁻¹⁶), confirming persistent level differences in national stocks among camels, cattle, goats, and sheep.

The time smooths are highly significant for all species (all p \< 2×10⁻¹⁶), showing that each livestock series follows a non-linear trajectory over 1981–2024. Cattle, goats, and sheep show substantial non-linearity (edf ≈ 7.82, 8.21, and 6.90, respectively), consistent with multi-phase dynamics rather than a single monotonic trend. Camels show a much simpler time structure (edf ≈ 1.0), consistent with a smoother long-run trajectory. In practical terms, the dominant signal in the data is “time structure” (multi-decadal changes and regime-like shifts), not short-term rainfall variability.

In contrast, the rainfall anomaly smooths (lags 0–2 years) are not statistically significant for most species–lag combinations at the 5% level. The only rainfall-related term close to conventional significance is the same-year anomaly effect for goats (p ≈ 0.055), while all other lag terms show weak evidence of association (p-values generally well above 0.1). This means that, at the national scale and after accounting for each species’ long-run trajectory, there is limited statistical evidence that year-to-year national rainfall anomalies (including 1–2 year delays) systematically explain additional variation in livestock stocks.

*Substantive meaning for the capstone question*: the model supports the conclusion that Somalia’s national livestock stocks exhibit strong, species-specific non-linear change over time, but it does not provide strong evidence that national mean rainfall anomalies are a major independent predictor of those national stock dynamics once time structure is included. This does not imply rainfall is unimportant biologically; rather, it indicates that national averaging may mask spatially localized climate–forage–herd mechanisms, and that livestock changes are also shaped by non-climatic drivers (conflict, markets, mobility constraints, disease, policy, and data-quality/estimation regimes). A clearer rainfall–livestock signal may emerge in regional analyses or in models that use spatial rainfall summaries aligned with pastoral systems rather than a single national mean.

```{r}
# 1) Ensure correct types in livestock table 
somalia_livestock_stocks_1981_2024 <- somalia_livestock_stocks_1981_2024 %>%
  mutate(
    year    = as.integer(year),     # year as integer
    value   = as.numeric(value),    # stocks as numeric
    species = as.factor(species)    # species as factor (grouping variable)
  )

# 2) Build national rainfall anomaly lags
rain_lags <- som_ts_mean %>%
  select(year, anomaly) %>%         # keep only what we need
  arrange(year) %>%                 # guarantee time order
  mutate(
    anom_l0 = anomaly,              # same-year anomaly (t)
    anom_l1 = lag(anomaly, 1),      # 1-year lag (t-1)
    anom_l2 = lag(anomaly, 2)       # 2-year lag (t-2)
  ) %>%
  select(year, anom_l0, anom_l1, anom_l2)

# 3) Merge livestock with rainfall lags by year
dat_gam <- somalia_livestock_stocks_1981_2024 %>%
  inner_join(rain_lags, by = "year") %>%   # match years
  filter(!is.na(anom_l2)) %>%              # drop early years where lags are missing
  arrange(species, year) %>%               # sort within species
  mutate(
    log_value = log(value)                 # log-transform stocks for stability
  )

# 4) Fit GAM: species-specific non-linear time trend + species-specific non-linear rainfall effects (lags 0–2)
gam_fit <- mgcv::gam(
  log_value ~ species +
    s(year,    by = species, k = 10) +
    s(anom_l0, by = species, k = 6) +
    s(anom_l1, by = species, k = 6) +
    s(anom_l2, by = species, k = 6),
  data   = dat_gam,
  method = "REML"
)

# 5) Print model results 
summary(gam_fit)
# 6) Diagnostics (checks residual patterns, k adequacy, etc.)
par(mfrow = c(2, 2))
gam.check(gam_fit)
par(mfrow = c(1, 1))

# 7) Partial effect plots (non-linear shapes by species and by lag)
#    These are the key plots to interpret: how livestock responds across anomaly values (wet/dry)
par(mfrow = c(3, 2))
plot(gam_fit, pages = 1, shade = TRUE, seWithMean = TRUE)
par(mfrow = c(1, 1))

# 8) Observed vs fitted stocks (back-transform to original scale)
  dat_gam <- dat_gam %>%
  mutate(
    pred_log   = predict(gam_fit, newdata = dat_gam, type = "response"),
    pred_value = exp(pred_log))


#plot
ggplot(dat_gam, aes(x = year)) +
  geom_line(aes(y = value,     color = "Observed"),   linewidth = 0.8) +
  geom_line(aes(y = pred_value, color = "GAM fitted"), linewidth = 0.8, linetype = "dotted") +
  facet_wrap(~ species, scales = "free_y", ncol = 2) +
  scale_color_manual(values = c("Observed" = "steelblue", "GAM fitted" = "darkred"),name = NULL ) +
  scale_y_continuous(labels = label_comma()) +
  labs(
    title = "Observed vs GAM-fitted livestock stocks (FAOSTAT) with rainfall anomaly lags (CHIRPS)",
    subtitle = "Non-linear time trend + rainfall anomaly lags (0–2 years), estimated separately by species",
    x = "Year",
    y = "Stocks (heads)",
    caption = "Dashed red lines represent the model's ability to capture non-linear trends"

  ) +
  
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(size = 10),
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text = element_text(face = "bold"),
    legend.position = "right",
    panel.grid.minor = element_blank()
  )
```


```{r}
ggplot(dat_gam, aes(x = anom_l0, y = log_value, color = species)) +

  geom_point(alpha = 0.45, size = 2) +

  geom_smooth(method = "gam", se = TRUE, linewidth = 1.1) +

  geom_vline(xintercept = 0, linetype = "dashed", color = "black") +

  scale_color_viridis_d(name = "Livestock species") +

  labs(
    title = "Relationship Between Rainfall Anomalies and Livestock Stocks in Somalia",
    
    subtitle = "Scatter plot of national rainfall anomalies (CHIRPS v3) versus log-transformed livestock stocks (FAOSTAT)\nEach point represents a species–year observation (1983–2024); smooth curves show GAM-estimated trends",
    
    x = "Rainfall anomaly (mm) relative to the 1981–2024 mean",
    
    y = "Log-transformed livestock stocks (log(heads))",
    
    caption = "Data sources: CHIRPS v3 rainfall data (Climate Hazards Center) and FAOSTAT livestock stocks.\nLines represent GAM smooth fits with confidence intervals."
  ) +

  theme_minimal(base_size = 13) +

  theme(
    plot.title = element_text(face = "bold", size = 15),
    plot.subtitle = element_text(size = 11),
    legend.position = "right",
    legend.title = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )
```

# Discussion

## Discussion

The results indicate that Somalia's rainfall patterns are dominated by strong 
interannual variability rather than a consistent long-term trend. Both the 
OLS and Mann–Kendall tests show little evidence of a monotonic rainfall trend 
over the study period.

Livestock population trends differ substantially by species. Camels show a 
significant long-term increase, while goats show a strong decline and cattle 
remain relatively stable. These differences suggest that livestock adaptation 
strategies may vary across species.

The GAM analysis indicates that rainfall anomalies alone do not strongly explain 
national livestock stock changes once long-term time dynamics are included. 
This suggests that additional drivers such as conflict, market changes, 
disease outbreaks, and pastoral mobility constraints may play an important role.

## Limitations

Several limitations should be considered when interpreting these results.

First, livestock data from FAOSTAT are national aggregates and may include 
estimated values, especially during periods of limited data reporting. 
Second, rainfall was analyzed at the national level, which may mask 
important local climate variability affecting pastoral systems. 
Third, other drivers such as conflict, land-use change, livestock disease, 
and market dynamics were not explicitly modeled in this analysis.

Future research could integrate regional livestock data and spatially explicit
climate variables to better capture localized climate–livestock relationships.

## Policy Implications

Improving climate resilience in Somalia's pastoral sector requires 
better integration of climate monitoring and livestock management systems. 
Early-warning systems, drought preparedness strategies, and improved 
rangeland management can help reduce vulnerability to rainfall variability.

In addition, strengthening livestock data collection at regional levels 
would allow policymakers to better understand how climate variability 
affects pastoral systems across different ecological zones.


# Conclusion

This study analyzed long-term rainfall variability and livestock population 
dynamics in Somalia using CHIRPS climate data and FAOSTAT livestock statistics 
from 1981 to 2024.

Results show that rainfall variability is characterized primarily by strong 
year-to-year fluctuations rather than a long-term trend. Livestock populations 
exhibit species-specific trajectories, with camels increasing significantly, 
goats declining, and cattle remaining relatively stable.

Statistical modeling suggests that national rainfall anomalies alone do not 
strongly explain livestock population changes once long-term time patterns 
are considered. These findings highlight the importance of considering both 
climate and socio-economic drivers when analyzing pastoral system dynamics.

Understanding these dynamics can support more effective climate adaptation 
strategies for Somalia’s pastoral sector.


# References

Climate Hazards Center. (2024). *Climate Hazards Group InfraRed Precipitation with Station Data (CHIRPS), Version 3*.University of California, Santa Barbara.  
https://www.chc.ucsb.edu/data/chirps 

Food and Agriculture Organization of the United Nations. (2024).  
*FAOSTAT statistical database: Livestock stocks*.  
https://www.fao.org/faostat/en/

Global Administrative Areas (GADM). (2023).  
*GADM database of global administrative boundaries (Version 4.1)*.  
https://gadm.org/

Wood, S. N. (2017). *Generalized additive models: An introduction with R* (2nd ed.).  
Chapman and Hall/CRC.

Hijmans, R. J. (2023). *terra: Spatial data analysis (R package)*.  
https://cran.r-project.org/package=terra

Pebesma, E. (2018). *Simple features for R: Standardized support for spatial vector data*.  
The R Journal, 10(1), 439–446.

Wickham, H., et al. (2019). *Welcome to the tidyverse*. Journal of Open Source Software, 4(43), 1686.
  